From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-31 paper.Rmd
WIP

---
title: "From noise to knowledge: how randomness generates novel phenomena and information"
author:
  - name: "Carl Boettiger"
    affiliation: a, 1
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"

layout: preprint
abstract: |
  In this paper, I provide a brief synthesis of ecological research in stochastic phenomena. Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].  Stochastic phenomena can suggest new ways of inferring process from pattern, and thus spark more dialog between theory and empirical perspectives that best advances the field as a whole. I highlight a few compelling examples, while observing that the study of stochastic phenomena are only beginning to make this translation into empirical inference.  There are rich opportunities at this interface in the years ahead.

@@

lineno: true

bibliography: refs.bib
output: rticles::elsevier_article
---

@@

<!-- Regarding simulation examples -->
In this spirit, each of the examples illustrated here is coupled with concise and efficient R code in the appendix to both reproduce and further explore the simulations.  While examples of noisy phenomena in natural systems are also crucial, matching the pattern does not necessarily mean we have also matched the process behind the data.  The examples cited from the literature here offer some of the best empirical evidence of these phenomena, but I encourage readers to also experiment with the example code provided.  Simulations permit poking and prodding of empirical investigation quickly and with little mathematical formalism in the way.  In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulas. This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017]. A copy of this appendix is maintained at <https://github.com/cboettig/noise-phenomena>, and bug reports, suggestions or help requests are welcome through the issue tracker.  A second appendix provides more mathematical details for interested readers.  With these in place, we are ready to begin our tour through three major areas of noisy phenomena: Origins of noise, emergent phenomena, and noise-driven inference.

@@

What is noise? Where does it come from?  It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.  Simply put, *noise is all that we leave out of our model*.  Frequently, this additional noise is tacked on to the end of a model, as if as an afterthought,

$$X_{t+1} = f(X_t) + \xi_t$$

with the assertion that $\xi$ is some normally distributed noise term reflecting the combined effects of any large array of possible factors that also influence the value of our state, $X$.  While not always wrong, this hand-waving over-simplification is increasingly replaced with a more ecologically and mechanistically grounded description for the origins of noise.  With the exception of quantum mechanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent property of nature but rather an emergent property of many events occurring at a different scale.  Coins and dice obey the physical laws of motion; only variation in how they are tossed (i.e. something complex we do not usually model) creates randomness.  The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise. Statistical mechanics tells us that temperature is a measure of the average speed of individual molecules in a gas -- any given molecule will experience a range of possible speeds but by itself has no notion of temperature.  Likewise, stochasticity enters our ecological models because our quantities of interest, such as the density of a population, is but an average over a lower-level, discrete process of individual births and deaths. Stochasticity that emerges directly from discrete individual events that underlie the process described by our model are considered *intrinsic stochasticity*, commonly referred to as *demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to population level descriptions. An important aspect of this demographic stochasicity is that it scales with the square root of the "system size," so that for very large populations the resulting fluctuations are relatively small. On the other hand, *extrinsic stochasticity*, usually referred to as *environmental stochasticity* in the ecological literature, arises in processes that are not described by our model, but merely impact the value of some parameter of our model; for instance, fluctuations in temperature may drive variation in a death rate, or fluctuations in rainfall drive variation in carrying capacity.

<!--
Early ecological models were almost always formulated at the level of the population, described in terms of differential or difference equations reflecting the available tools of the time.  The advent of fast and inexpensive computing power 
-->

@Black2010 provides an excellent review of the importance of intrinsic stochasticity and the evolution

Consider Levins' patch model, where $n$ individuals compete for a finite number of suitable habitats $N$.  Individuals die a constant rate $e$, and produce offspring at a constant rate $c$ who then have a probability of colonizing an open patch that is simply proportional to the fraction of available patches, $1 - n/N$.

Approximations of the so-called master equation, van Kampen system size approximation [@vanKampen2007] described in @Black2010, (rooted in the theorems of [@Kurtz1970; @Kurtz1971]),

$$\frac{\mathrm{d} \langle n \rangle }{dt}  = b(\langle n \rangle) - d(\langle n \rangle)$$
$$\frac{\mathrm{d} \sigma^2 }{dt}  = 2 \left( \frac{\partial b(n) }{\partial n} - \frac{\partial d(n) }{\partial n}\right) \sigma^2 + b(n) + d(n) $$

<!-- 
SDE notation

$$\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t$$

An SDE. Note the separation of the differential elements, which reveal that this is actually shorthand for an integral equation:

$$X_{t+s} - X{t}  = \int_t^{t+s} \mu(X_{\tau}, \tau) \textrm{d}\tau + int_t^{t+s} \sigma(X_{\tau}, \tau) \textrm{d}B\tau$$

The first integral is the ordinary Lebesgue integral, while the second is an It\^o integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process). The It\^o calculus follows different rules from the classic calculus, largely due to the fact that an It\^o integral has non-zero quadratic variation. (Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 ~ dx$).

While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.  Obviously events such as births and deaths are themselves the results of mechanistic processes, just as fluctuations in rainfall and temperature are the result of deterministic laws of physics and not fundamentally random forces.  In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.  
<!-- Our state variables don't "exist" except as averages of a lower level process -->

**Demographic stochasticity**.  Demographic stochasticity refers to fluctuations in population sizes or densities that arise from the fundamentally discrete nature of individual birth and death events.  Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a fundamentally lower-level mechanistic process.  We summarize the myriad lower-level processes that mechanistically lead to the event of a 'birth' in the population as a probability: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.  Assuming each of these events are independent, this is a Poisson process.

@@

<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual. Show figure with small and large N  -->

[@Black2012]

**Environmental stochasticity**.  Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes operating at the same scale that we are not explicitly modelling.

@@

@@

Colored noise:
- [@Ripa1999]
- [@Marshall2015]

<!-- Identify as beyond the scope of this paper -->
<!-- non-temporal stochasticity -->

Several additional sources of stochasticity or variation more generally fall outside these categories demographic and environmental noise.  While a detailed treatment is beyond the scope of this paper, I enumerate these sources to disambiguate such phenomena from the intrinsic and extrinsic noise sources that comprise the focus here.

**Other forms of stochasticity**.  Stochasticity frequently enters our models in other ways as well.  For instance, most population-level models treat all individuals within the population, or at least within a particular age or stage class, as identical.  *Individual heterogeneity* in relevant traits (growth rates, dispersal ability) can be represented probabilistically as a distribution rather than a fixed value, making it function as another form of random variation in a model.  *Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit model, or as spatially varying environmental stochasticity in a spatially explicit model.  As with other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial variation and other sources of noise can drive patterns including the potential for persistence, coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017].

**Observational error**: Observational error acts much like another form of stochasticity, arising from
sampling effects or limitations of instruments, etc. Yet unlike the sources described above, this 
uncertainty does not impact the actual state of our system, and thus does not give rise to any
additional phenomena or information about that process, though it can obscure it [@Knape2011; @deValpine2002].

Other forms of uncertainty common in models are not stochastic phenomena.  This includes any form of 
**reducible uncertainty**: In constructing and estimating mathematical models of ecological processes, 
we encounter uncertainty over our parameter estimates and possible choice of models [@Ellison2004].  In some ways, this
uncertainty can act like the stochasticity described above: for instance, both introduce variation into
any forecast of future outcomes. Yet unlike stochasticity, this uncertainty is reducible: with additional
observations, we can narrow the uncertainty over parameters or candidate models, while stochasticity remains an
inherent source of variation. We will this not consider model and parameter uncertainty, as they are not 
sources of noise.

**Chaos**: Deterministic chaos [@Desharnais2001] is another mechanism of introducing significant variation into a simple model. Chaos can a look a lot like noise: indeed, it can be observationally indistinguishable from a stochastic process [@Werndl2009].  Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Costantino1997] the prevalence of chaotic dynamics as a driving factor in most ecological systems remains unclear. After all, the great variation of natural populations is hardly surprising: the undeniable influences of variation created by all those things left out of our models, all that individual detail from intrinsic, lower level processes and external environmental processes that becomes aggregated in our simple models all but begs for a stochastic description.

@@

Stochasticity can give rise to a wide array of interesting phenomena that are often at odds with our intuition. Such behavior often arises through an interaction between stochastic and non-linear elements of a system, both hallmark traits of ecological processes.  In this section, I highlight several examples that have shown up frequently in the literature and prove sufficiently tractable to provide a simple intuition for the mechanisms involved.

<!-- Say something about how small noise; particularly demographic noise which is usually small, seems like it should thus be negligible,

**Noise need not be large to be important**.
-->

**Persistence and coexistence**.  A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species. Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guarenteed to eventually go extinct under demographic noise.  While extinction is certainly a phenonomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself. Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.  In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Schreiber2017].

<!--
[@Schreiber2008] complex life histories
- Species coexistence  ([@Hart2016])
-->

**Bet hedging**

**Inflation**

Stochasticity can

<!-- two concurrent flat lines makes for a poor figure -->

@@

Oscillations arise through the process of stochastic resonance.  The damped oscillations seen in Figure 1a are analgous to a child sitting still on a swing: friction slowly damps the magnitude of each successive oscillation, while the period of pendulum remains unchanged.  Driving the pendulum through perturbations timed to match that natural period can quickly create sustained oscillations with large magnitudes, but such perfect timing is not required.  In the model above, white noise effectively drives the pendulum at all frequencies, though at very low magnitude, since the noise is small relative to the scale of the dynamics.  However, frequencies that most closely match the natural period of the damped oscillator are amplified, resonating with the natural oscillations.  This process is sufficient to permit even small noise to drive sustained oscillations in the model indefinitely.

The power spectrum (the squared norm of the Fourier transform, a measure of what frequencies are present; see, e.g. @Black2012) provides a convenient way to visualize the effects of these oscillations.  The power spectrum  for a deterministic limit cycle would create a perfect sinusoidal oscillation and a resulting power spectrum of a delta spike at that single frequency.  The power spectrum for the quasi-cycle instead shows a distribution of frequencies, centered at the the resonant frequency of the damped oscillator.

Early work by @Blarer1999 identified quasi-cycles as a possible explanation for oscillations and periodic outbreaks in pest ecosystems.  @Bjornstad2004 analyzes the potential for such stochastic effects to drive cyclic behavior in population abundances of Atlantic bluefin tuna and Atlantic Cod, while @Black2010 illustrates the use of power-spectrum approaches to examine the behavior of quasi-cycles in common disease outbreak models, focusing on a case where noise arises explicitly from demographic stochasticity, rather than generic environmental noise as in the model above and elsewhere.  @Pineda-Krch2007 compares patterns from these stochastically generated cycles to truly stable limit cycles they resemble.

Stochastic resonance highlights one way in which even small noise can play an important role in driving large scale patterns where we had previously only sought deterministic explanations. This example also illustrates the interplay between nonlinear dynamics and stochasticity that is the hallmark of stochastically driven phenomena.  A stable node with linear dynamics, such as an Ornstein-Uhlenbeck process, has no resonant frequency, and could not produce sustained oscillations (unless the environmental driver itself had a periodic pattern).

<!-- variable noise -->

<!--autocorrelated or red noise can also drive stochastic resonance -->

@@

<!--Figure 3: Noisy switch -->
```{r noisy_switch, fig.width=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}

@@

theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  mutate(growth = x * p$r * (1 - x / p$K), 
         consumption = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  mutate(potential = - cumsum(growth - consumption)) %>%
  gather(curve, y, -x, -potential)

@@

p2 <-  theory %>%
  ggplot(aes(x, y, col=curve)) +
  geom_line(lwd=1) + 
  scale_color_ptol() + 
  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.)))

@@

gridExtra::grid.arrange(p1,p2,p3, layout_matrix = rbind(c(1,1), c(2,3)))

#g$widths <- unit.pmax(g1$widths, g2$widths) # use the largest widths
#grid.newpage()
#grid.draw(g)
```

This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.  This behavior arises through the interaction of stochasticity and alternative stable state dynamics.  The simulation shown in Figure 3A is based on an herbivore consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics, to which I have simply introduced an additional environmental noise term:

$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption, } c(X_t)} + \xi_t,$$
where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$. Figure 3B shows separate curves corresponding the parts of the model describing vegetation growth and vegetation consumption, using the same parameters as shown in the simulation panel, 3A.  Equilibria exist whenever growth balances consumption.  For the parameters shown, the S-shaped curve of the Type-III consumption pattern crosses the quadratic curve of logistic growth at four separate locations.  When growth is larger than consumption before the crossing, and thus smaller after, the point is stable, otherwise it is unstable.  Consequently, we can see two stable points, in the neighborhood of $X_t \approx 1.2$ and $X_t \approx 0.5$, closely matching the average fluctuations seen in the simulation in 3A.  Stochastic fluctuations drive spontaneous shifts between these alternative stable states.

@@

**Further examples**

- @Read2007  Stochasticity generates an evolutionary instability for infectious disease 
- Spatial pattern formation: Turning patterns found in plankton-herbivore dynamics [@Levin1976]

@@

One of the most interesting aspects of noise is as a source of additional information.  Noise can act like a miniature perturbation experiment, constantly perturbing a system away from an equilibrium to reveal it's behavior elsewhere in state space.

- Measuring stability

**Early warning signals.**  The theory of early warning signals, or critical slowing down (CSD) [@Wissel1984; @Scheffer2009] is an instructive example

>>>>

---
title: "From noise to knowledge: how randomness generates novel phenomena and information"
author:
  - name: "Carl Boettiger"
    affiliation: a
    email: "cboettig@berkeley.edu"

address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  5p # review = doublespace, 3p = singlespace, 5p = two-column
#preamble: "\\usepackage[nomarkers]{endfloat}"

abstract: |
  In this paper, I provide a brief synthesis of ecological research in stochastic phenomena. Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].  Stochastic phenomena can suggest new ways of inferring process from pattern, and thus spark more dialog between theory and empirical perspectives that best advances the field as a whole. I highlight a few compelling examples, while observing that the study of stochastic phenomena are only beginning to make this translation into empirical inference.  There are rich opportunities at this interface in the years ahead.

@@

---

@@

<!-- Regarding simulation examples -->
In this spirit, each of the examples illustrated here is coupled with concise and efficient R code in the appendix to both reproduce and further explore the simulations.  While examples of noisy phenomena in natural systems are also crucial, matching the pattern does not necessarily mean we have also matched the process behind the data.  The examples cited from the literature here offer some of the best empirical evidence of these phenomena, but I encourage readers to also experiment with the example code provided.  Simulations permit poking and prodding of empirical investigation quickly and with little mathematical formalism in the way.  In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulas. This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017]; allowing the deductive models illustrated here to readily function as inductive models with parameters inferred from time-series data. A copy of this appendix is maintained at <https://github.com/cboettig/noise-phenomena>, and bug reports, suggestions or help requests are welcome through the issue tracker.  A second appendix provides more mathematical details for interested readers.  With these in place, we are ready to begin our tour through three major areas of noisy phenomena: Origins of noise, emergent phenomena, and noise-driven inference.

@@

What is noise? Where does it come from?  It is important to bear in mind that whenever stochasticity is introduced into a model,
it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
Simply put, *noise is all that we leave out of our model*.  Early quantiative ecological models first formulated at the population level
as differential or difference equations: deterministic models that seek to capture the essential features of a system while neglecting
any random component as mere noise [@Coulson2004]. Advances in both theoretical understanding and the advent of widely available and
efficient numerical methods have helped drive research towards increasingly more explicit descriptions of how stochastiticity enters
into quantiative models [@Black2012; @Ovaskainen2010].

With the exception of quantum mechanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent
property of nature but rather an emergent property of many events occurring at a different scale.  Coins and dice obey the physical
laws of motion; only variation in how they are tossed (i.e. something complex we do not usually model) creates randomness.  The term
"noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually
distinct blends together into a background of noise. Statistical mechanics tells us that temperature is a measure of the average speed
of individual molecules in a gas -- any given molecule will experience a range of possible speeds but by itself has no notion of temperature.
Likewise, stochasticity enters our ecological models because our quantities of interest, such as the density of a population, is but an
average over a lower-level, discrete process of individual births and deaths. Stochasticity that emerges directly from discrete individual
events that underlie the process described by our model are considered *intrinsic stochasticity*, commonly referred to as
*demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to
population level descriptions. An important aspect of this demographic stochasicity is that it scales with the square root of the
"system size," so that for very large populations the resulting fluctuations are relatively small. On the other hand, 
*extrinsic stochasticity*, usually referred to as *environmental stochasticity* in the ecological literature, arises in processes
that are not described by our model, but merely impact the value of some parameter of our model; for instance, fluctuations in
temperature may drive variation in a death rate, or fluctuations in rainfall drive variation in carrying capacity.

<!--
Motivate the presentation of the van Kampen result rather than a generic review.
-->

## Demographic stochasticity

Demographic stochasticity refers to fluctuations in population sizes or densities that arise from the fundamentally discrete nature of individual birth and death events.  Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a lower-level mechanistic process.  We summarize the myriad lower-level processes that mechanistically lead to the event of a 'birth' in the population as a probability: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.  Assuming each of these events are independent, this is a state-dependent Poisson process. The change in the probability of being in state $N$ is given by the sum over the ways to enter the state, minus the ways to leave the state: a simple expression of probability balance known as the "master equation."

It is possible to simulate such a process exactly using Gillespie's exact stochastic simulation algorithm [SSA; @Gillespie1977], which simulates
each individual birth or death as a separate event.
Consider Levins' patch model, where $n$ individuals compete for a finite number of suitable habitats $N$.
Individuals die a constant rate $e$, and produce offspring at a constant rate $c$ who then have a probability
of colonizing an open patch that is simply proportional to the fraction of available patches, $1 - n/N$. 
Figure 1 shows the results of two exact SSA simulations of this model with identical parameters except for the
total number of available sites, $N$.

The van Kampen system size expansion provides a convenient way to explicitly approximate the "macroscopic" (i.e. population level, see @Black2012) dynamics from an individual-based description of events in a Markov process [@vanKampen2007; based on @Kurtz1970; @Kurtz1971].  In addition to illustrating how common stochastic models (in particular, Guassian noise) arise from lower-level discriptions, this result will also provide us a useful framework for interpeting the behavior of stochastic models more generally.  In Appendix B I illustrate the derivation of the following equations for the mean and expected population sizes for a generic density dependent birth-death process:

\begin{align}
\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) \\
\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \frac{\partial b(x) }{\partial x} - \frac{\partial d(x) }{\partial x}\right) \sigma^2 + b(x) + d(x) \label{bd}
\end{align}

where $x = \langle n \rangle$, the average or expected population size of an ensemble of replicates.  We can solve these equations at steady-state by setting both equations zero; e.g. for our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.  Note that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.

@@

## Environmental stochasticity

Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes whose dynamics we are not explicitly modelling.  A population-level model

@@

[@Ovaskainen2010] Master equation, break-down of Fokker-Planck approximation under large noise, introduction of the WKB approximation

@@

Colored noise: [@Lawton]
- [@Ripa1999]
- [@Marshall2015]

### Other forms of stochasticity

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals within the population, or at least
within a particular age or stage class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically as a distribution rather than
a fixed value, making it function as another form of random variation in a model [@Coulson2001; @Schreiber2010].
*Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit
model, or as spatially varying environmental stochasticity in a spatially explicit model.  As with
other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial
variation and other sources of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017].
*Observational error*: Observational error acts much like another form of stochasticity, arising from
sampling effects or limitations of instruments, etc. Yet unlike the sources described above, this 
uncertainty does not impact the actual state of our system, and thus does not give rise to any
additional phenomena or information about that process, though it can obscure it [@Knape2011; @deValpine2002].

Other forms of uncertainty common in models are not stochastic phenomena.  This includes any form of 
*reducible uncertainty*: In constructing and estimating mathematical models of ecological processes, 
we encounter uncertainty over our parameter estimates and possible choice of models [@Ellison2004].  In some ways, this
uncertainty can act like the stochasticity described above: for instance, both introduce variation into
any forecast of future outcomes. Yet unlike stochasticity, this uncertainty is reducible: with additional
observations, we can narrow the uncertainty over parameters or candidate models, while stochasticity remains an
inherent source of variation. We will this not consider model and parameter uncertainty, as they are not 
sources of noise.  *Deterministic chaos* [@Desharnais2001] is another mechanism of introducing significant
variation into a simple model. Chaos can a look a lot like noise: indeed, it can be observationally
indistinguishable from a stochastic process [@Ellner2005; @Werndl2009].  Despite substantial interest in previous
decades and the relative ease of encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations
[@Costantino1997] the prevalence of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations is hardly surprising: the
undeniable influences of variation created by all those things left out of our models, all that
individual detail from intrinsic, lower level processes and external environmental processes that
becomes aggregated in our simple models all but begs for a stochastic description.

@@

Stochasticity can give rise to a wide array of interesting phenomena that are often at odds with our intuition. As these examples will illustrate, such behavior arises through an interaction between stochastic and non-linear elements of a system, underscoring the importance of explicitly including stochasticity in ecological models [@Coulson2004; @Higgins1997].  Importantly, the noise itself does not have to be large to give rise to these phenomena. In this section, I highlight several examples that have shown up frequently in the literature and prove sufficiently tractable to provide a simple intuition for the mechanisms involved.

<!-- 
Noise is "tacked on" in these models... contrasts to previous section
-->

## Persistence and coexistence

A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species. Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guarenteed to eventually go extinct under demographic noise.  While extinction is certainly a phenonomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself. Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.  In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].

Insufficient space left for any treatment of these...

**Bet hedging**
**Inflation**

@@

Oscillations arise through the process of stochastic resonance.  The damped oscillations seen in Figure 1a are analgous to a child sitting still on a swing: friction slowly damps the magnitude of each successive oscillation, while the period of pendulum remains unchanged.  Driving the pendulum through perturbations timed to match that natural period can quickly create sustained oscillations with large magnitudes, but such perfect timing is not required.  In the model above, white noise effectively drives the pendulum at all frequencies, though at very low magnitude, since the noise is small relative to the scale of the dynamics.  However, frequencies that most closely match the natural period of the damped oscillator are amplified, resonating with the natural oscillations.  This process is sufficient to permit even small noise to drive sustained oscillations in the model indefinitely.  The power spectrum (the squared norm of the Fourier transform, a measure of what frequencies are present; see, e.g. @Black2012) provides a convenient way to visualize the effects of these oscillations.  The power spectrum  for a deterministic limit cycle would create a perfect sinusoidal oscillation and a resulting power spectrum of a delta spike at that single frequency.  The power spectrum for the quasi-cycle instead shows a distribution of frequencies, centered at the the resonant frequency of the damped oscillator.

Early work by @Blarer1999 identified quasi-cycles as a possible explanation for oscillations and periodic outbreaks in pest ecosystems.  @Bjornstad2004 analyzes the potential for such stochastic effects to drive cyclic behavior in population abundances of Atlantic bluefin tuna and Atlantic Cod, while @Black2010 illustrates the use of power-spectrum approaches to examine the behavior of quasi-cycles in common disease outbreak models, focusing on a case where noise arises explicitly from demographic stochasticity, rather than generic environmental noise as in the model above and elsewhere.  @Pineda-Krch2007 compare patterns from these stochastically generated cycles to stable limit cycles they resemble.  Stochastic resonance highlights one way in which even small noise can play an important role in driving large scale patterns where we had previously only sought deterministic explanations. This example also illustrates the interplay between nonlinear dynamics and stochasticity that is the hallmark of stochastically driven phenomena.  A stable node with linear dynamics, such as an Ornstein-Uhlenbeck process, has no resonant frequency, and could not produce sustained oscillations (unless the environmental driver itself had a periodic pattern).

<!-- See additional citations in Coulson2004 and others that mention stochastic resonance -->

@@

<!--Figure 3: Noisy switch -->
```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}

@@

theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  mutate(g = x * p$r * (1 - x / p$K), 
         c = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  mutate(potential = - cumsum(g - c)) %>%
  gather(curve, y, -x, -potential)

@@

p2 <-  theory %>%
  ggplot(aes(x, y, col=curve)) +
  geom_line(lwd=1) + 
  scale_color_ptol() + 
  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.))) +
  theme(legend.position="top")

@@

gridExtra::grid.arrange(p1,p2,p3, layout_matrix = rbind(c(1,1), c(2,3)))
```

This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.  This behavior arises through the interaction of stochasticity and alternative stable state dynamics.  The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.  The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:

$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{ growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{consumption, } c(X_t)} + \xi_t,$$
where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$. Figure 3B shows separate curves corresponding the parts of the model describing vegetation growth and vegetation consumption, using the same parameters as shown in the simulation panel, 3A.  Equilibria exist whenever growth balances consumption.  For the parameters shown, the S-shaped curve of the Type-III consumption pattern crosses the quadratic curve of logistic growth at four separate locations.  When growth is larger than consumption before the crossing, and thus smaller after, the point is stable, otherwise it is unstable.  Consequently, we can see two stable points, in the neighborhood of $X_t \approx 1.2$ and $X_t \approx 0.5$, closely matching the average fluctuations seen in the simulation in 3A.  Stochastic fluctuations drive spontaneous shifts between these alternative stable states.

@@

<!--
**Further examples**
- @Read2007  Stochasticity generates an evolutionary instability for infectious disease 
- Spatial pattern formation: Turning patterns found in plankton-herbivore dynamics [@Levin1976]

@@

One of the most interesting aspects of noise is as a source of additional information.  This can be realized in a variety of ways. For instance, in deriving expressions for the variance created by demographic noise above, we observed that different partitions of a logistic growth model can lead to the same macroscopic dynamics (same values for $b$ - $d$), while having different variance (due to different values for $b$+$d$).  This suggests that the variation observed in such a population could resolve potentially help resolve how dynamics are partitioned between individual birth and death rates (though in practice one would have to first account for any contribution of environmental noise).

<!-- fixme need more background? -->
